---
title: "How to avoid counfounding in Hedwig's Study? It depends on which DAG we assume"
author: "Joseph Bulbulia"
date: "12/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(#fig.path="figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                      fig.width= 10,
                      fig.height=8,
                      tidy = "styler")
```
                      
## Question

Does psychopathy lead to promotion?



<!-- Ask what variables we need to condition on. To assess the effect of psychopathy on success we must stratify on male. -->
<!-- ```{r} -->
<!-- dagify( -->
<!--   Success ~ Psycho + Age + Male + Edu + JobTime + Years, -->
<!--   Psycho ~ Male, -->
<!--   JobTime ~ Age + Years, -->
<!--   Edu ~ Years, -->
<!--   labels = c( -->
<!--     "Psycho" = "Psychopathy", -->
<!--     "Success" = "Success", -->
<!--     "Age" = "Age", -->
<!--     "Male" = "Male", -->
<!--     "Edu" = "Edu", -->
<!--     "JobTime" = "JobTime", -->
<!--     "Years" = "Years" -->
<!--   ) -->
<!-- ) %>% -->
<!--   ggdag_adjustment_set( -->
<!--     text = FALSE, -->
<!--     use_labels = "label", -->
<!--     shadow = TRUE, -->
<!--     exposure =  "Psycho", -->
<!--     outcome =   "Success", -->
<!--   ) + theme_dag_blank() -->
<!-- ``` -->
<!-- ```{r} -->
<!-- dagify( -->
<!--   Success ~ Psycho + Age + Male + Edu + JobTime + Years, -->
<!--   Psycho ~ Male, -->
<!--   JobTime ~ Age + Years, -->
<!--   Edu ~ Years, -->
<!--   labels = c( -->
<!--     "Psycho" = "Psychopathy", -->
<!--     "Success" = "Success", -->
<!--     "Age" = "Age", -->
<!--     "Male" = "Male", -->
<!--     "Edu" = "Edu", -->
<!--     "JobTime" = "JobTime", -->
<!--     "Years" = "Years" -->
<!--   ) -->
<!-- ) %>% -->
<!--   ggdag_dseparated( -->
<!--     from = "Psycho", -->
<!--     to = "Success", -->
<!--     controlling_for = c( "Years",  "Age",  "Male",  "Edu" , "JobTime"),  -->
<!--     text = FALSE, -->
<!--     use_labels = "label" -->
<!--   ) + theme_dag_blank() -->

<!-- ``` -->


<!-- ```{r} -->
<!-- dagify( -->
<!--   Success ~ Psycho + Male + Edu + JobTime + Years, -->
<!--   Psycho ~ Male, -->
<!--   JobTime ~  Years, -->
<!--   Edu ~ Years, -->
<!--   labels = c( -->
<!--     "Psycho" = "Psychopathy", -->
<!--     "Success" = "Success", -->
<!--     "Age" = "Age", -->
<!--     "Male" = "Male", -->
<!--     "Edu" = "Edu", -->
<!--     "JobTime" = "JobTime", -->
<!--     "Years" = "Years" -->
<!--   ) -->
<!-- ) %>% -->
<!--   ggdag_adjustment_set( -->
<!--     text = FALSE, -->
<!--     use_labels = "label", -->
<!--     shadow = TRUE, -->
<!--     exposure =  "Psycho", -->
<!--     outcome =   "Success", -->
<!--   ) + theme_dag_blank() -->
<!-- ``` -->


# Which variables do we need to "control" for confounds?


Our interest is in causation not predictive accuracy.  

For this reason we must take care when including control ariables that obtain unbiasted estimate for the effect of psychopathic personality on promotion.

First we need to find the adjustment set for assessing whether psychopathy predicts promotion. 

```{r}
library("dagitty")
library("ggdag")
dag2<-dagify(
  Promotion ~  Success0 +  JOBTIME1 + Edu1 + Psycho +  MALE + U,
  Success0 ~ JOBTIME0 + Edu0 + MALE + Psycho,
  JOBTIME1 ~ JOBTIME0 + Edu1 + Psycho, # impulsivity? 
  JOBTIME0 ~  Edu0 + Psycho,
  Psycho ~ MALE, # unmeasured variable afffects both success and education 
  Success0 ~ U,# unmeasured variable afffects both success and education, call it "priviledge"
  Edu0 ~ U,
  Edu1 ~ U,
  Edu1 ~ Edu0,
    labels = c(
    "Promotion" = "Promotion",
    "Success0" = "Success0",
    "JOBTIME1" = "JOBTIME1",
    "Edu1" = "Edu1",
    "Psycho" = "Psycho",
    "MALE" = "MALE",
     "U" = "Unmeasured",
    "JOBTIME0" = "JOBTIME0",
    "Edu0" = "Edu0"
    )
  ) 

ggdag_adjustment_set(dag2,
    text = FALSE,
    use_labels = "label",
    shadow = TRUE,
    exposure =  "Psycho",
    outcome =   "Promotion",
  ) + theme_dag_blank()
```

We observe that to assess the question of which variables to include to assess the effect of pyschopathic personality on promotion we need to condition on Male.  Note that we don not need to include more than Male to assess whether psychopathy predicts promotion, according to this causal DAG. 

We might ask, would including the other variables introduce confounding? After all, we might also want better prediction. The answer is yes. Including the other measures opens confounding from the unmeasured source of success. 


```{r}
ggdag_dseparated(dag2,
    from = "Psycho",
    to = "Promotion",
    controlling_for = c( "Success0" , "U", "MALE", "Edu1","Edu0","JOBTIME0","JOBTIME1"), 
    text = FALSE,
    use_labels = "label",
    collider_lines = TRUE,
  ) + theme_dag_blank()
```


What if we inlcude only MALE, as our DAG instructed us? The collider confounding closes. 

```{r}
ggdag_dseparated(dag2,
    from = "Psycho",
    to = "Promotion",
    controlling_for = c( "MALE"), 
    text = FALSE,
    collider_lines = TRUE,
    use_labels = "label") + theme_dag_blank()

```

What if we were to include only a subset of "control" variables?  Perhaps adjusting for level of success at baseline is important. Perhaps people at the bottom of the heap will not be promoted as quickly as those at the top, especially if they are psychopaths. 


```{r}
ggdag_dseparated(dag2,
    from = "Psycho",
    to = "Promotion",
    controlling_for = c( "MALE", "Success0"), 
    text = FALSE,
    collider_lines = TRUE,
    use_labels = "label") + theme_dag_blank()

```


We need to close the door.  If we include only MALE we do that.
 
```{r}
ggdag_dseparated(dag2,
    from = "Psycho",
    to = "Promotion",
    controlling_for = c( "MALE"), 
    text = FALSE,
    collider_lines = TRUE,
    use_labels = "label") + theme_dag_blank()

```


## Summmary
As of today, my sense is that to understand whether Psychopathy causes promotion, we would be wise to consider only conditioning only on MALE. Doing so closes the backdoor between unmeasured influences on promotion that might also affect previous success that operate differently for those who identify as male and those who do not. 

It is a worry that small changes to a DAG will make big differences to which causal feaures of the worlk we think our results are describing. 

I don't know how close the DAGS I've presented here are to reality. I use them to illustrate the need to proceed with caution when introducing control variables. 


<!-- NOTE: -->
<!-- Causal inference gets much more complicated in a heirarchical modelling setting such as ours. -->

<!-- https://bpspsychub.onlinelibrary.wiley.com/doi/abs/10.1111/bmsp.12217 -->



