---
title: "Statistical Rethinking | Chapter 5"
author: "Joseph Bulbulia"
date: "10/01/2021"

output:
  ioslides_presentation:
    incremental: no
   # mathjax: local
    self_contained: false
    widescreen: no
    smaller: yes
    font_adjustment: -5
    css: extra.css
editor_options: 
    markdown: 
    wrap: 72
---


```{r global_options,  include = FALSE}
knitr::opts_chunk$set(message=FALSE, 
                      warning=FALSE,
                      colmlapse =TRUE,
                      echo=TRUE)
                      #results="hide", 
                     # fig.width = 10,
                     # fig.height=8)
# read libraries
```
```{r sourcelibraries, include=FALSE}
source("libs.R")
#source("funs.R") # custom functions for analysis
```

## Multiple regression, why? 

1. "control" i.e. confounding
2. complex causation (mediation)
3. interactions -- importance of one var depends on another var


## Load Data
```{r}
# load data and copy
library(rethinking)
data(WaffleDivorce)
d <- WaffleDivorce

# standardize variables
d$D <- standardize( d$Divorce )
d$M <- standardize( d$Marriage )
d$A <- standardize( d$MedianAgeMarriage )
```

## Single regressions

```{r cache = TRUE}
m5.1 <- quap(
  alist(
    D ~ dnorm( mu , sigma ) ,
    mu <- a + bA * A ,
    a ~ dnorm( 0 , 0.2 ) ,
    bA ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dexp( 1 )
    ) , data = d )
precis(m5.1)
```


## Single regression with marriage rate


```{r cache =TRUE}
m5.2 <- quap(
  alist(
    D ~ dnorm( mu , sigma ) ,
    mu <- a + bM * M ,
    a ~ dnorm( 0 , 0.2 ) ,
    bM ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dexp( 1 )
) , data = d )

precis(m5.2)
```

## Thinks before you regress!
Only one predictor has a causal effect.

```{r}
# possible dags
library(dagitty)
library(ggdag)
dagify(
  D  ~ A,
  M ~ A
) %>%
  ggdag() +
  theme_dag_blank()
```
 
## Another possible DAG

```{r}
dagify(D ~ A + M,
       M ~ A) %>% 
  ggdag() +
  theme_dag_blank()
```

## Implied conditional independence

```{r}
library(dagitty)
DMA_dag2 <- dagitty('dag{ D <- A -> M }')
impliedConditionalIndependencies( DMA_dag2 )
```

D $\perp$ M $|$ A


## Multiple regression

```{r echo = FALSE}
m5.3 <- quap(
  alist(
    D ~ dnorm( mu , sigma ) ,
    mu <- a + bM*M + bA*A ,
    a ~ dnorm( 0 , 0.2 ) ,
    bM ~ dnorm( 0 , 0.5 ) ,
    bA ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dexp( 1 )
    ) , data = d )
plot( coeftab(m5.1,m5.2, m5.3), par=c("bA","bM") )
```

## Simulate Divorce example

```{r cache = TRUE, echo = FALSE}
library(brms)
library(sjPlot)
N <- 500 # number of simulated States
age <- rnorm( N )# sim A
mar <- rnorm( N , -age ) # sim A -> M”
div <- rnorm( N ,  age ) # sim A -> D”
#create data frame
nd<-cbind(age,mar,div)
#run models in brms
b1<- brms::brm(div ~ age, data = nd)
b2 <- brms::brm(div ~ mar, data = nd)
b3 <- brms::brm(div ~ age + mar, data = nd)

bayesplot::mcmc_intervals(b3,pars = c("b_age", "b_mar"))
```

## Masked Relationships

```{r, echo = FALSE}
library(ggdag)
library(patchwork)
p1 <- dagify(
  N  ~ M,
  K ~ N + M
) %>%
  ggdag() +
  theme_dag_blank()

p2 <- dagify(
  M  ~ N,
  K ~ N + M
) %>%
  ggdag() +
  theme_dag_blank()


p3<- dagify(
  K ~ N + M,
  N ~ U,
  M ~ U
) %>%
  ggdag() +
  theme_dag_blank()
p1 + p2 + p3 + plot_annotation(title = "DAGS are equivalent, can't infer relationships", tag_levels = 'a')
```



## Quote

> "Whenever an article or book says something like "we found no difference" or "no effect," this usually means that some parameter was not significantly different from zero, and so the authors adopted zero as the estimate. This is both illogical and extremely common." -- SR, p. 241



## Next week collider bias

```{r}
ggdag_collider_triangle(
  x = NULL,
  y = NULL,
  m = NULL,
  x_y_associated = FALSE,
  edge_type = "link_arc",
  node_size = 16,
  text_size = 3.88,
  label_size = text_size,
  text_col = "white",
  label_col = text_col,
  node = TRUE,
  stylized = FALSE,
  text = TRUE,
  use_labels = NULL
)

```




