---
title: "Statistical Rethinking | Notes for Chapter 4"
author: "Joseph Bulbulia"
date: "09/01/2021"
output:
  ioslides_presentation:
    incremental: no
   # mathjax: local
    self_contained: false
    widescreen: yes
    smaller: yes
    font_adjustment: -5
    css: extra.css
editor_options: 
    markdown: 
    wrap: 72
---


```{r global_options,  include = FALSE}
knitr::opts_chunk$set(message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE)
                      #results="hide", 
                     # fig.width = 10,
                     # fig.height=8)
# read libraries
```
```{r sourcelibraries, include=FALSE}
source("libs.R")
#source("funs.R") # custom functions for analysis
```


## Recap
- Chapter 3  Generate samples from "quap" and usuing them to create data summaries.  

## Overview

- introduction to the gaussian distribution 

- a language for describing models

- using this langauge to write a model
  - writing priors 
  - writing a model (using priors)

- linear regression * 

- nonlinear regression 

<!-- <div class="notes"> -->
<!-- *our focus today = linear regression -->
<!-- </div> -->

## language for describing models

$$
y_i \sim Normal ( \mu_i, \sigma)\\
\mu_i \sim \beta x_i \\
\beta \sim Normal(0,10) \\
\sigma \sim Exponentional (1) \\
x_i \sim Normal (0, 1)
$$




## What should we expect height to be?  

In most cases, priors are specified independently for each parameter, which amounts to assuming: 

$$\Pr(\mu,\sigma) = \Pr(\mu)\Pr(\sigma)$$

i.e. we will have priors for *both* the mean and the standard deviation 

## Prior for height
```{r}
curve( dnorm( x , 178 , 20 ) , from=100 , to=250 )
```

## Prior for Standard Deviation \ (contrained to be postive)

```{r}
curve( dunif( x , 0 , 50 ) , from=-10 , to=60 )
```

## Let's generate samples

```{r}
sample_mu <- rnorm( 1e4 , 178 , 20 ) # prior on avg height
sample_sigma <- runif( 1e4 , 0 , 50 ) # prior on standard deviation
prior_h <- rnorm( 1e4 , sample_mu , sample_sigma ) # generate the prior for height
dens( prior_h )
```

## Here are some thoughtless samples
```{r}
sample_mu <- rnorm( 1e4 , 150 , 1000 ) # prior on avg height
sample_sigma <- runif( 1e4 , 0 , 1000 ) # prior on standard deviation
prior_h <- rnorm( 1e4 , sample_mu , sample_sigma ) # generate the prior for height
dens( prior_h )
```
<div class="notes">
We don't think height should be like this.
</div>



## Load Howell's height data

data
```{r howell, include=TRUE}
# data from rethinking
data( Howell1 )
# assig data to object `d`
d <- Howell1
# filter ages greater than 18
d2 <- d[ d$age >= 18 , ]
```


## Graph of heigh ~ weight
```{r howellgraph}
plot( d2$height ~ d2$weight )
```


## translation of model into code
<div class="columns-2">
$$
h_i \sim Normal(\mu, \sigma) 
$$
$$
\mu \sim  Normal(170,20)
$$
$$
\sigma \sim Uniform (0,50)
$$




-  `height ~ dnorm( mu, sigma ) ` 
-  `mu ~ dnorm( 178, 20 )`
-  `sigma ~ dunif( 0, 50 )`

## Model
<div class="columns-2">
Model
```{r}
flist <- alist(
  height ~ dnorm( mu , sigma ) ,
  mu ~ dnorm( 178 , 20 ) ,
  sigma ~ dunif( 0 , 50 )
)
# model
m4.1<- quap( flist , data=d2 )
```

Summary
```{r}
rethinking::precis( m4.1)
```

## Try with initalised starting values
<div class="columns-2">

Model with start
```{r}
start <- list(
  mu=mean(d2$height),
  sigma=sd(d2$height)
)
m4.1<- quap( flist , data=d2 )
```


Summary
```{r}
rethinking::precis( m4.1)
```

## Linear Regression 

-  `height ~ dnorm( mu, sigma )  [likelihood]` 
-  `mu <- alpha + beta(x_i - xbar) [so called 'linear" model]`
-  `alpha ~ dnorm( 178, 20 ) [alpha prior]`
-  `beta ~ Normal ( 0 , 10) [beta prior]`
-  `sigma ~ dunif( 0, 50 ) [sima prior]`

## Simulate prior{ .tiny}
```{r code plot of priors}
set.seed(2971)
N <- 100 # 100 lines
a <- rnorm(N , 178 , 20)
b <- rnorm(N , 0 , 10)
xbar <- mean(d2$weight)
plot(NULL ,  xlim = range(d2$weight) , ylim = c(-100, 400) , xlab = "weight" ,  ylab = "height" )
abline(h = 0 , lty = 2); abline(h = 272 , lty = 1 , lwd = 0.5); mtext("b ~ dnorm(0,10)")
# plot lines
for (i in 1:N)
  curve(  a[i] + b[i] * (x - xbar) ,
    from = min(d2$weight) , to = max(d2$weight) ,add = TRUE , col = col.alpha("black", 0.2) )
```

## Graph 
```{r plot of priors, echo = FALSE}
set.seed(2971)
N <- 100 # 100 lines
a <- rnorm(N , 178 , 20)
b <- rnorm(N , 0 , 10)
xbar <- mean(d2$weight)
plot(NULL ,  xlim = range(d2$weight) , ylim = c(-100, 400) , xlab = "weight" ,  ylab = "height" )
abline(h = 0 , lty = 2); abline(h = 272 , lty = 1 , lwd = 0.5); mtext("b ~ dnorm(0,10)")
# plot lines
for (i in 1:N)
  curve(  a[i] + b[i] * (x - xbar) ,
    from = min(d2$weight) , to = max(d2$weight) ,add = TRUE , col = col.alpha("black", 0.2) )
```

## Try same with sensible priors
<div class="columns-2">

```{r}
b <- rlnorm( 1e4 , 0 ,1)
dens( b , xlim=c(0,5) , adj=0.1 )
```
```{r}
dev.off()
#set up plot
plot( NULL , xlim = range(d2$weight) , ylim = c(-100, 400) , xlab = "weight" , ylab = "height")
abline(h = 0 , lty = 2)
abline(h = 272 , lty = 1 , lwd = 0.5)
mtext("b ~ dnorm(0,10)")
# plot lines
b <- rlnorm( 1e4 , 0 ,1)
for (i in 1:N)
  curve( a[i] + b[i] * (x - xbar) ,
         from = min(d2$weight) , to = max(d2$weight) , add = TRUE , col = col.alpha("black", 0.2))
```

## Now we fit the regression model

```{r  fig.show='hide'}
d <- Howell1
d2 <- d[ d$age >= 18 , ]
# define the average weight, x-bar
xbar <- mean(d2$weight)
# fit model
m4.3 <- quap(
  alist(
  height ~ dnorm(mu , sigma) ,
  mu <- a + b * (weight - xbar) ,
  a ~ dnorm(178 , 20) ,
  b ~ dlnorm(0 , 1) ,
  sigma ~ dunif(0 , 50)
) ,
data = d2)
precis(m4.3)
```

## plot

```{r cache = TRUE}
plot(height ~ weight , data = d2 , col = rangi2)
post <- extract.samples(m4.3)
a_map <- mean(post$a)
b_map <- mean(post$b)
curve(a_map + b_map * (x - xbar) , add = TRUE)
```

## Uncertainty Graph | extract samples from the model 

```{r cache = TRUE, fig.show='hide'}
N <- 10
dN <- d2[1:N ,]
mN <- quap(alist(
  height ~ dnorm(mu , sigma) ,
  mu <- a + b * (weight - mean(weight)) ,
  a ~ dnorm(178 , 20) ,
  b ~ dlnorm(0 , 1) ,
  sigma ~ dunif(0 , 50)
) ,
data = dN)
#extract 20 samples from the posterior
post <- extract.samples(mN , n = 20)

plot( dN$weight ,  dN$height , xlim = range(d2$weight) , ylim = range(d2$height) , col = rangi2 , xlab = "weight" , ylab = "height")
mtext(concat("N = ", N))
for (i in 1:20)
  curve(post$a[i] + post$b[i] * (x - mean(dN$weight)) ,
        col = col.alpha("black", 0.3) ,
        add = TRUE)
```

## Plot

```{r cache = TRUE, echo = FALSE}
N <- 10
dN <- d2[1:N ,]
mN <- quap(alist(
  height ~ dnorm(mu , sigma) ,
  mu <- a + b * (weight - mean(weight)) ,
  a ~ dnorm(178 , 20) ,
  b ~ dlnorm(0 , 1) ,
  sigma ~ dunif(0 , 50)
) ,
data = dN)
#extract 20 samples from the posterior
post <- extract.samples(mN , n = 20)

plot( dN$weight ,  dN$height , xlim = range(d2$weight) , ylim = range(d2$height) , col = rangi2 , xlab = "weight" , ylab = "height")
mtext(concat("N = ", N))
for (i in 1:20)
  curve(post$a[i] + post$b[i] * (x - mean(dN$weight)) ,
        col = col.alpha("black", 0.3) ,
        add = TRUE)
```


## Graph of expected mean at a certain value

Mean at certain value
```{r cache = TRUE}
post <- extract.samples( m4.3 )
mu_at_50 <- post$a + post$b * (50 - xbar)
dens (mu_at_50, xlab = "mu|weight = 50", col = "CadetBlue4") # for base R colors see: http://www.sthda.com/english/wiki/colors-in-r
```

## Calculate HPDI
```{r}
PI(mu_at_50, prob = 0.89)
```

## Shortcut for computing means for each case: `link`

```{r cache = TRUE}
mu_0 <- link(m4.3)  # just those in dataset
weight.seq <-seq(from = 25, to = 70, by = 1) # hypothetical population
mu <- link(m4.3, data = data.frame(weight = weight.seq))# use type = 'n' to hide raw data
plot (height ~ weight, d2, type = "n")
for ( i in 1:100)  points( weight.seq, mu[i,],  pch = 16,  col = col.alpha(rangi2, 0.1))
```


## summarize the distribution of mu

```{r cache = TRUE}
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.89 )
# fading out points to make line and interval more visible
plot( height ~ weight , data=d2 , col=col.alpha(rangi2,0.5) )

# plot the MAP line, aka the mean mu for each weight
lines( weight.seq , mu.mean )

# plot a shaded region for 89% PI
shade( mu.PI , weight.seq )
```




## Note we can simulate heights from a population

```{r fig.show='hide'}
sim.height <- sim( m4.3 , data=list(weight=weight.seq) , n=1e4 )
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )
#plot raw data
plot( height ~ weight , d2 , col=col.alpha(rangi2,0.5) )
for ( i in 1:100)  points( weight.seq, mu[i,],  pch = 16,  col = col.alpha(rangi2, 0.1))
# draw HPDI region for line
shade( mu.PI  , weight.seq)
# draw PI region for simulated heights
shade( height.PI , weight.seq )
```

## Graph
```{r echo = FALSE}
sim.height <- sim( m4.3 , data=list(weight=weight.seq) , n=1e4 )
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )
#plot raw data
plot( height ~ weight , d2 , col=col.alpha(rangi2,0.5) )
for ( i in 1:100)  points( weight.seq, mu[i,],  pch = 16,  col = col.alpha(rangi2, 0.1))
# draw HPDI region for line
shade( mu.PI  , weight.seq)
# draw PI region for simulated heights
shade( height.PI , weight.seq )
```


## Interation 

<div class="columns-2">
$$
h_i \sim Normal(\mu_i, \sigma) 
$$
$$
\mu_i = \alpha + \beta_1 x_i + \beta_2 x_i^2
$$

$$
\alpha \sim Normal(178,20)
$$
$$
\beta_1 \sim LogNormal(0,1)
$$
$$
\beta_2 \sim Normal(0,1)
$$

$$
\mu \sim  Normal(170,20)
$$
$$
\sigma \sim Uniform (0,50)
$$


- `height ~ dnorm( mu, sigma ) `
- `mu = a + b1 * weight_s + b2 * weight_s^2`
-  `a ~ dnorm( 178, 20 )`
- `b ~ dlnorm( 0, 1 )`
- `b2 ~ norm( 0, 1)`
- `sigma ~ dunif( 0, 50 )`

## Model

```{r}
data(Howell1)
d <- Howell1 # note using all age data here not just over 18
d$weight_s = scale(d$weight)
d$weight_s2 = d$weight_s ^ 2

m4.5 <- quap(
  alist(
    height ~ dnorm(mu , sigma) ,
    mu <- a + b1 * weight_s + b2 * weight_s2 ,
    a ~ dnorm(178 , 20) ,
    b1 ~ dlnorm(0 , 1) ,
    b2 ~ dnorm(0 , 1) ,
    sigma ~ dunif(0 , 50)
  ) ,
  data = d
)
```

## summary

```{r}
precis(m4.5)
```

## plot
```{r echo = FALSE}
weight.seq <- seq( from=-2.2 , to=2 , length.out=30 )
pred_dat <- list( weight_s=weight.seq , weight_s2=weight.seq^2 )
mu <- link( m4.5 , data=pred_dat )
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI , prob=0.89 )
sim.height <- sim( m4.5 , data=pred_dat )
height.PI <- apply( sim.height , 2 , PI , prob=0.89 )

plot( height ~ weight_s , d , col=col.alpha(rangi2,0.5) )
lines( weight.seq , mu.mean )
shade( mu.PI , weight.seq )
shade( height.PI , weight.seq )
```


## We'll hold off on splines for now, but skim chapter

TBC