---
title: "Statistical Rethinking | Notes for Chapter 4"
author: "Joseph Bulbulia"
date: "09/01/2021"
output:
  ioslides_presentation:
    incremental: no
   # mathjax: local
    self_contained: false
    widescreen: yes
    smaller: yes
    font_adjustment: -1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r global_options,  include = FALSE}
knitr::opts_chunk$set(message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=FALSE)
                      #results="hide", 
                     # fig.width = 10,
                     # fig.height=8)
# read libraries
```
```{r sourcelibraries, include=FALSE}
source("libs.R")
#source("funs.R") # custom functions for analysis
```
## Recap
- Chapter 3  Generate samples from "quap" and usuing them to create data summaries.  

## Overview

- introduction to the gaussian distribution 

- a language for describing models

- using this langauge to write a model
  - writing priors 
  - writing a model (using priors)

- linear regression * 

- nonlinear regression 

<div class="notes">
* our focus today = linear regression
</div>

## language for describing models

$$y_i \sim Normal ( mu_i, \sigma)\\
\mu_i \sim \beta x_i \\
\beta \sim Normal(0,10) \\
\sigma \sim Exponentional (1) \\
x_i \sim Normal (0, 1)
$$



## Howell's height data

```{r howell, include=FALSE}
# data from rethinking
data( Howell1 )

# assig data to object `d`
d <- Howell1

# filter ages greater than 18
d2 <- d[ d$age >= 18 , ]
```

# Langauge