---
title: "sr_chapter_1_notes"
author: "Joseph Bulbulia"
date: "01/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Highlights from Chapter 1 in Statistical Rethinking

A few key points in Statistical Rethinking Chapter 1

## Golems and Bayes
* RM points out that statistical tests are "golems" -- little robots -- that will destroy your life, and Prague.
* These robots work well in some settings, but they are fragile. They fail in spectacular ways.
* Bayesian data analysis is a golem. However, it has many advantages, one of which is recovering intuitive probabilities.

## Statitical Models, Process Models, and Hypotheses

  + Models: how does the world work, e.g. a process model of evolution
  + Hypotheses: predictions that models make on observations: 
  + Statistical model: evaluation of a hypothesis: do these observations make a model more probable?  
  + It is not just statistical tests that may go wrong: 
    * need to understand the relationship between a process model and its observations
    * need to evaluate (and re-evaluate) a statistical model (Note, a hypotheses might be continuous)
    * measurement matters 
  
##Practical Advice

Storm clouds on the horizon: causal inference requires causal models: we need to be careful about what we include in our regressions: confounds are lurking.

Key tools at our disposal:

 1. Bayesian data analysis
 2. Model comparison
 3. Multilevel models
 4. Graphical causal models


# Questions

* What does RM think is wrong about classical statistical instruction? 
* What does RM think is OK about classical statistical instruction? 
* What is the advantage of Bayesian data analysis
* How do multilevel models deal with variation. 


  
# Practicalities (not in the book)


```{r}
# comments will help you, and others, following your statistical reasoning
# here we load data
data(cars)

# I like to inspect any data that I load
str(cars)
head(cars)

# here's a basic model
m <- lm( dist ~ speed , data=cars )
summary(m)
```


Let's obtain some extra packages that will make reporting useful for you. 
```{r}
# tip of the day; many packages make reporting easy. For example the sjPlot package

# you need the devtools library. in Rstudio click "packages --> install --> devtools"

if (!require(devtools)) {
  install.packages("devtools")
}
if (!require(sjPlot)) {
 devtools::install_github("strengejacke/sjPlot")
}

if (!require(ggeffects)) {
 install.packages("ggeffects")
}

if (!require(ggplot2)) {
 install.packages("ggplot2")
}

```

Here's a nicely formatted table

```{r table_model_m}
# table for the model
tab_model(m)
```

Here's a coefficient plot for the model, notice how I have customised it

```{r coefficientplot_model_m}
sjPlot::plot_model(m) + ggtitle("Relationship of speed to distance")
```


Here's a prediction plot

```{r prediction_plot}
p <- ggpredict(m)
plot(p) 
```


# Code for installing the rethinking package


```{r install_rethinking}
# function for installing dependencies
ipak <- function(pkg){
new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
sapply(pkg, require, character.only = TRUE)
}

# usage
packages <- c("coda", "plyr", "mvtnorm", "scales", "dagitty")
ipak(packages)

# next install rethinking
if (!require(rethinking)) {
  devtools::install_github("rmcelreath/rethinking")
}

```










